---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pipr

<!-- badges: start -->
[![R-CMD-check](https://github.com/worldbank/pipr/workflows/R-CMD-check/badge.svg)](https://github.com/worldbank/pipr/actions?workflow=R-CMD-check)
[![test-coverage](https://github.com/worldbank/pipr/workflows/test-coverage/badge.svg)](https://github.com/worldbank/pipr/actions)
[![pkgdown](https://github.com/worldbank/pipr/workflows/pkgdown/badge.svg)](https://github.com/worldbank/pipr/actions)
[![Codecov test coverage](https://codecov.io/gh/worldbank/pipr/branch/main/graph/badge.svg)](https://app.codecov.io/gh/worldbank/pipr?branch=main)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![R-CMD-check](https://github.com/worldbank/pipr/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/worldbank/pipr/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/worldbank/pipr/graph/badge.svg)](https://app.codecov.io/gh/worldbank/pipr)
<!-- badges: end -->

The `pipr` package allows R users to compute poverty and inequality indicators 
for more than 160 countries and regions from the World Bank’s database of 
household surveys. It does so by accessing the Poverty and Inequality Platform 
(PIP) API. PIP is a computational tool that allows users to estimate poverty 
rates for regions, sets of countries or individual countries, over time and at 
any poverty line.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("worldbank/pipr")
```

## Example

This is a basic example that shows how to retrieve some key poverty and inequity statistics. 

### Retrieve statistics

```{r example, warning=FALSE, message=FALSE}
library(dplyr)
library(pipr)

df <- get_stats(country = "ALB")
glimpse(df)
```

### Access data dictionary
```{r}
get_aux("dictionary")
```


## Citation

To cite package `pipr` in publications use:

```
  Tony Fujs, Aleksander Eilertsen, Ronak Shah and R. Andrés Castañeda (2022). pipr: Client for the PIP
  API. https://github.com/worldbank/pipr, https://worldbank.github.io/pipr/.
```

A BibTeX entry for LaTeX users is

```
  @Manual{,
    title = {pipr: Client for the PIP API},
    author = {Tony Fujs and Aleksander Eilertsen and Ronak Shah and R. Andrés Castañeda},
    year = {2022},
    note = {https://github.com/worldbank/pipr,https://worldbank.github.io/pipr/},
  }
```

