---
title: "pipr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pipr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
NOT_CRAN <- identical(tolower(Sys.getenv("NOT_CRAN")), "true")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  purl = NOT_CRAN
)
```

```{r setup}
library(pipr)
```

The `pipr` package allows to easily query the [PIP API](https://pip.worldbank.org/api) from R.

Here are a few examples to get you started:

## Using default options

By default, `get_stats()` returns all available  

- countries  
- reporting levels (national, urban, rural)  
- for all available survey years  
- at the 2.15 USD a day poverty line (in 2017 PPP)  

For the most recent data available.

```{r message=FALSE, warning=FALSE, eval=NOT_CRAN}
get_stats()
```

## Basic options

### Filter by country

```{r, warning=FALSE, message=FALSE, eval=NOT_CRAN}
# Specify ONE country
get_stats(country = "ALB")

# Specify MULTIPLE countries
get_stats(country = c("ALB", "CHN"))
```

### Filter by year

By default, `get_stats` will only return data if a survey is available for the requested __country__ and the requested __year__.

```{r, warning=FALSE, message=FALSE, eval=NOT_CRAN}
# Survey year available
get_stats(country = "ALB", year = 2012)         # single year
get_stats(country = "ALB", year = c(1996,2012)) # multiple years

# Survey year NOT available - Empty response
get_stats(country = "ALB", year = 2000)

# Most Recent Value available
get_stats(country = "ALB", year = "MRV")

```

### Modify the poverty line
```{r, warning=FALSE, message=FALSE, eval=NOT_CRAN}
get_stats(country = "ALB", povline = 3.2)
```

## Other features

### Get estimates when survey year is not available

The `fill_gaps` argument triggers the interpolation / extrapolation of poverty 
estimates when surveys are not available for a given year.

```{r, warning=FALSE, message=FALSE, eval=NOT_CRAN}
# fill_gaps = FALSE (default)
get_stats(country = "HTI")
# fill_gaps = TRUE
get_stats(country = "HTI", fill_gaps = TRUE)
```

### Compute aggregates

`get_stats()` can also be used to compute aggregate welfare statistics for sub-regions.
Only World Bank regions are currently available, but more sub-regional groupings
will become available in the future.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
# World Bank aggregate
get_stats(country = "all", subgroup = "wb_regions")
# List of World Bank regions
get_regions()
```
